---
title: "Trends in Book Genres Over the Past Century"
author: "Oleksandra Krykun"
output:
  html_document: default
  pdf_document: default
---

## Introduction

In this report, I present an analysis of the "Best Books Ever" dataset from Kaggle, available at [this link](https://www.kaggle.com/datasets/pooriamst/best-books-ever-dataset). 

The focus of this analysis is to explore **trends in book genres** over the period from 1916 to 2017 and its **distribution**. This involved extensive data cleaning and engineering to prepare the dataset for visualization.

---

## Importing Libraries and Data

Before diving into the analysis, we need to load the necessary libraries and the dataset. The following libraries are used for data manipulation, visualization, and interactive plotting:

```{r}
# Install necessary packages (if not already installed)

# install.packages("plotly")
# install.packages("dplyr")
# install.packages("tidyverse")
# install.packages("ggplot2")
# install.packages("lubridate")

# Load libraries
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))

# Load the dataset
books <- read.csv("C:\\Users\\kriku\\Documents\\Data Projets\\book_analysis_in_R\\books_1.Best_Books_Ever.csv")
```

## Data Preparation

**Step 1: Filtering and Cleaning Genre Data**

To analyze genre trends, we first extract relevant columns (bookId, genres, and firstPublishDate), clean the data, and transform the firstPublishDate into a usable format. We also ensure that rows with missing or invalid data are removed.

```{r}
suppressWarnings(
genre_data <- books %>%
  select(bookId, genres, firstPublishDate) %>%
  filter(!is.na(genres) & genres != "[]" & !is.na(firstPublishDate) & firstPublishDate != "") %>% 
  separate_rows(genres, sep = ", ")  %>% 
  mutate(genres = gsub("[[:punct:]]", "", genres),
         firstPublishDate = substr(firstPublishDate, 7, 8),            # select the year part of the date
         firstPublishDate = ifelse(as.numeric(firstPublishDate) > 15,  # convert to full year
                                   paste0("19", firstPublishDate),  
                                   paste0("20", firstPublishDate)),
         firstPublishDate = paste0(firstPublishDate, "-01-01"),
         firstPublishDate = year(ymd(firstPublishDate)))  %>% 
  filter(!is.na(firstPublishDate)) %>% 
  arrange(desc(firstPublishDate))  
)

# view(genre_data)
```

**Step 2: Selecting Top 5 Genres per Year**

To simplify the visualization, we focus on the top 5 most popular genres for each year.

```{r}
df_top5 <- genre_data %>%
  group_by(firstPublishDate, genres) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(firstPublishDate) %>%
  slice_max(order_by = count, n = 5) %>% 
  ungroup() 

# view(df_top5)
```

**Step 3: Counting Total Books Published per Year**

We also create a separate dataframe to track the total number of books published each year.

```{r}
year_count <- genre_data %>% 
  select(bookId, firstPublishDate) %>% 
  distinct(bookId, .keep_all = TRUE) %>% 
  count(firstPublishDate) %>% 
  suppressWarnings(
  mutate(firstPublishDate = paste0(firstPublishDate, "-01-01"),
         firstPublishDate = year(ymd(firstPublishDate)))
  )

# view(year_count)
```

## Visualizing Genre Trends Over Time

We combine the top 5 genres and the total number of books published into a single plot to visualize trends over time.

```{r, fig.width=10, fig.height=6}
ggplot() +
  geom_area(data = df_top5, aes(x = firstPublishDate, y = count, fill = genres), 
            position = "stack", alpha = 0.8) +
  geom_line(data = year_count, aes(x = firstPublishDate, y = n), 
            color = "black", linewidth = 1) +
  labs(title = "Top 5 Genres and Total Number of Books Published Over Time",
       x = "Year",
       y = "Number of Books",
       fill = "Genre") +
  theme_minimal()
```

### Distribution of the **book's genre**

```{r}
# Create the dataframe of genres counts
top_genres <- genre_data %>%
  separate_rows(genres, sep = ", ") %>%
  mutate(genres = gsub("[[:punct:]]", "", genres)) %>%
  count(genres, sort = TRUE) %>%
  slice_max(n, n = 30)

# Define custom colors for the pie chart
colors <- c('#1f77b4', '#ff7f0e', '#2ca02c', '#d62728', '#9467bd',
            '#8c564b', '#e377c2', '#7f7f7f', '#bcbd22', '#17becf',
            '#aec7e8', '#ffbb78', '#98df8a', '#ff9896', '#c5b0d5',
            '#c49c94', '#f7b6d2', '#c7c7c7', '#dbdb8d', '#9edae5')

# Create the enhanced pie chart
plot_ly(data = top_genres, 
        labels = ~genres, 
        values = ~n, 
        type = "pie",
        textinfo = "label+percent", 
        hoverinfo = "label+percent+value", 
        marker = list(colors = colors, line = list(color = '#FFFFFF', width = 1)),
        textposition = "inside", 
        insidetextfont = list(color = '#FFFFFF'), hole = 0.3) %>% 
  layout(title = "Top 25 Genres by Book Count", 
         showlegend = TRUE, 
         legend = list(font = list(size = 12)),
         margin = list(t = 50, b = 50, l = 50, r = 50)
         ) 
```

## Conclusion

This analysis reveals key trends in book genres and its distribution over the past century:

- **Rising Popularity**: Genres such as Young Adult, Fantasy, Fiction, and Romance have seen increasing popularity

- **Declining Genres**: In contrast, genres like Novels, Historical Fiction and 20th Century literature have experienced a notable decline, suggesting a shift in literary tastes over time.

- **Dominant Genres**: Fiction (14.9%) and Fantasy (6.23%) remain the most dominant genres, consistently leading in book publications. Classics, Novels, and Young Adult also hold significant shares, as illustrated by the area plot.

- **Overall Growth**: The steady rise in total book publications over the years underscores the expanding literary landscape and the growing diversity of genres available to readers.

_linkedin_: https://www.linkedin.com/in/oleksandra-krykun-0b45552a1/